Geospatial - Assignment 1

### 2025-01-28 hw1/submission_notebook.Rmd
Final optimizations on markdown file

In the following notebook we replicate the charts for all 5 papers. Contentwise, the code follows always the same structure. We first load the data, then we create the chart and finally display it.

@@

While the authors produced the same plot for three different years — 1950, 1980, and 2010 — due to a lack of data, we were only able to reproduce the plot for the most recent year, i.e., 2010 and with flaws for 1960 (i.e., we're missing around 13 mln people of the overall population in 1960 due to data compilation issues of the earlier census approaches of the IBGE such as non-digitalized records - [see here](https://ipeagit.github.io/censobr/reference/read_population.html#-census) for more information). Specifically, on the IBGE website, we found data for 2009, while the censobr package provides the 2010 data. It also provides census data for 1980 and 1960. However, for 1980, it seems that the population weights are missing. Therefore, we decided to focus on the 2010 and 1960 data.

@@

We reproduce the same chart for 2010, as the authors did. Due to data non-availability, we rely on a further chart for 1960, instead of 1950. The 1960 chart resembles the 1950 chart of the authors surprisingly well, despite the aforementioned data issues. Lastly, we omit 1980 data, since the population weights were missing in the censobr dataset and we were not able to infer them accurately.

@@

# Paper 4

@@

@@

@@

# Paper 5

@@

@@

@@

@@

![Figure x]()

>>>>

In the following notebook we replicate the charts for all 5 papers. Contentwise, the code follows always the same structure. We first load the data, secondly manipulate it, then we create the chart and finally display it.

@@

While the authors produced the same plot for three different years — 1950, 1980, and 2010 — due to a lack of data, we were only able to reproduce the plot for the most recent year, i.e., 2010 and with flaws for 1960 (i.e., we're missing around 13 mln people of the overall population in 1960 due to data compilation issues of the earlier census approaches of the IBGE such as non-digitalized records - [see here](https://ipeagit.github.io/censobr/reference/read_population.html#-census) for more information).

On the IBGE website, we found data for 2009, while the censobr package provides the 2010 data. It also provides census data for 1980 and 1960. However, for 1980, it seems that the population weights are missing. Therefore, we decided to focus on the 2010 and 1960 data.

@@

We reproduce the same chart for 2010, as the authors did. Due to data non-availability, we rely on a further chart for 1960, instead of 1950. The 1960 chart resembles the 1950 chart of the authors surprisingly well, despite the aforementioned data issues. The Northern regions being significantly different might be explained by the censobr authors who inter alia state that for 1960:
>"Information is missing for the states of the former Northern Region, Maranhão, Piaui, Guanabara, Santa Catarina, and Espírito Santo".

Lastly, we omit 1980 data, since the population weights were missing in the censobr dataset and we were not able to infer them accurately.

@@

# [Paper 4](https://economics.mit.edu/sites/default/files/publications/Catastrophe_Risk_and_Settlement_Location.pdf)

We are looking for a shapefile or dataset specifically containing road network data for Vietnam. We inspected data from [GADM](https://gadm.org/data.html#google_vignette) but none of the layers in the file appear to contain road network data. All layers are polygons representing administrative boundaries.

@@

```{r}
# Define the base path
path <- "data/paper4"

# We inspect data from Humdata:
st_layers(file.path(path, "ROADS/vnm_rdsl_2015_OSM.shp"))
# Read the shapefile into an sf object
vietnam_roads <- st_read(file.path(path, "ROADS/vnm_rdsl_2015_OSM.shp"))

# Get all the unique values from the column 'type'
unique_values <- unique(vietnam_roads$type)
print(unique_values)
```

Although the data is from 2015, whereas the paper displays roads from 2010, we decide to use this data as road networks.

```{r}
# We inspect data from SimpleMaps:
st_layers(file.path(path, "VIETNAM/"))

vietnam_map <- st_read(file.path(path, "VIETNAM", layer = "vn.shp"))
```

We decide to move forward with this data as Vietnam boundary.

@@

We use the [author's documentation](https://www.openicpsr.org/openicpsr/project/207641/version/V1/view;jsessionid=3EBB67FC7920B2090C0D74C349E99F93?path=/openicpsr/207641/fcr:versions/V1/Public-replication-files/Replication_README.pdf&type=file) to understand how he categorised the types of roads.

```{r}
# Categorise roads into group by trying to replicate the author's logic
vietnam_roads <- vietnam_roads %>%
  mutate(
    road_category = case_when(
      type %in% c("motorway", "motorway_link") ~ "Freeway",
      type %in% c("trunk", "trunk_link") ~ "Dual Carriageway",
      type %in% c("secondary") ~ "Major Roads",
      type %in% c("primary", "primary_link", "secondary_link", "tertiary", "tertiary_link", "residential", "living_street") ~ "Minor Roads",
      TRUE ~ "Other Roads"
    )
  )
```

@@

# [Paper 5](https://web.stanford.edu/~memorten/ewExternalFiles/Morten_Oliveira_Brasilia.pdf)

The author's replication package is accessible, so we move forward with it's [code and data](https://www.openicpsr.org/openicpsr/project/183316/version/V1/view).

@@

```{r}
path <- ("data/paper5/")

# Load the shapefile containing state boundaries for Brazil (1940 projection)
states <- st_read(file.path(path, "GIS_data/uf1940/uf1940_prj.shp"))
# Simplify the state boundaries to reduce file size while maintaining shape integrity
states_simple <- rmapshaper::ms_simplify(states, keep = 0.01, keep_shapes = TRUE)
```

@@

Since the authors' description is very comprehensive, we proceed also with their data manipulation approach.

```{r}
# Define the year for the highways data
year <- 2000

# Construct the file path for the highways shapefile dynamically based on the year
file_name <- paste0("/", year, "/highways_", year, "_prj.shp")

# Load the highways shapefile for the specified year
all_highways <- st_read(file.path(path, "GIS_data/roads", file_name))
# Simplify the highways data
all_highways_simple <- rmapshaper::ms_simplify(all_highways, keep = 0.01, keep_shapes = TRUE)

# Load and simplify the minimum spanning tree (MST) shapefile for pie distribution
mst_pie <- st_read(file.path(path, "GIS_data/mst/mst_pie_prj.shp"))
mst_pie_simple <- rmapshaper::ms_simplify(mst_pie, keep = 0.01, keep_shapes = TRUE)

# Load and simplify MST data excluding pies for Rio region
mst_all_rio <- st_read(file.path(path, "GIS_data/mst/mst_all_rio_prj.shp"))
mst_all_rio_simple <- rmapshaper::ms_simplify(mst_all_rio, keep = 0.01, keep_shapes = TRUE)

# Load shapefile for capital cities and extract coordinates for labeling
capital_cities <- st_read(file.path(path, "GIS_data/cities/brazil_capital_cities_prj.shp"))
cities_xy <- cbind(capital_cities, st_coordinates(capital_cities))
```

@@

Lastly, we also use their plotting approach:

```{r}
# Create a plot showing MST and highway networks in Brazil
ggplot() +
  # Add state boundaries as the base layer with light grey outlines
  geom_sf(data = states_simple, fill = "white", color = "grey90") +
  # Add the simplified MST data with a unique linetype and color
  geom_sf(data = mst_pie_simple, size = .6, linetype = "11", aes(color = "Minimum spanning tree"), show.legend = "line") +
  # Add non-radial highways with a dashed line style
  geom_sf(data = all_highways_simple %>% filter(dm_anlys_p == 1 & dm_radial == 0), size = .3, linetype = "dashed", aes(color = "Non-radial highways (2000)"), show.legend = "line") +
  # Add radial highways with a solid line style
  geom_sf(data = all_highways_simple %>% filter(dm_anlys_p == 1 & dm_radial == 1), size = 0.6, aes(color = "Radial highways (2000)"), show.legend = "line") +
  # Minimal theme for simplicity
  theme_minimal() +
  # Overlay capital cities and label them with their names
  geom_point(data = cities_xy, aes(x = X, y = Y)) +
  geom_text_repel(data = cities_xy, aes(x = X, y = Y, label = CITY_NAME)) +
  # Add a custom legend for line types and colors
  labs(color = " ") +
  scale_color_manual(
    values = c("#777676", "#868686", "#565555"),
    guide = guide_legend(override.aes = list(linetype = c("11", "dashed", "solid")))
  ) +
  # Remove axis lines, ticks, and text for a clean map appearance
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
```

@@

![Figure 1](data/paper5/original_plot.png)
